---
title: "Julia Frederick Coding Exercise"
output:
  html_document:
    df_print: paged
---
## Loading and checking data

```{r}
library(dslabs) #loading packages
data("gapminder") #loading data file
help("gapminder") #see what the data file contains
```
The gapminder dataset includes health and income outcomes for 185 countries from 1960 to 2016. Has two associated vectors, *oecd* and *opec*.

```{r}
str(gapminder) #shows the different variables, their class, a subset of the values associated, and how many lvls(if app)
summary(gapminder) # gives summary information for each variable 
                    #(if numeric-mean/median/etc; if character gives the number contained per level)
class(gapminder) # checks what type of object gapminder is -- "data.frame"
```


## Procesing data
Creating a dataframe with only countries in the African continent
```{r}
africadata <- subset(gapminder, continent == "Africa") # subsetting all data associated with the contient "Africa"
str(africadata) # checking the data included in this df, 2907 obs of 9 variables
summary(africadata) # summary information of all the variables
```

Breaking down the African dataset into smaller subsets
```{r}
#creating a new data frame, and naming the columns
lifecycle <- data.frame(infant_mortality = africadata$infant_mortality, life_expectancy = africadata$life_expectancy)
#creating a new data frame, and naming the columns
poplife <- data.frame(population = africadata$population, life_expectancy = africadata$life_expectancy)

str(lifecycle) #2907 obs of 2 variables
str(poplife) #2907 obs of 2 variables
summary(lifecycle) #averages of infant mortality and life expectancy
summary(poplife) #averages of population and life expectancy
```

## Plotting
Plot life expectancy as a function of infant mortality
```{r}
#plot(x value, y value, x-axis label, y-axis label, main title)
plot(lifecycle$infant_mortality, lifecycle$life_expectancy, xlab="Infant Mortality",
     ylab="Life Expectancy", main="Life expectancy as a function of infant mortality")
```
A negative correlation between infant mortality and life expectancy


Plot life expectancy as a function of populaiton size
```{r}
#plot(x value, y value, x-axis label, y-axis label, main title)
plot(log(poplife$population), poplife$life_expectancy, xlab="Log of Population Size", 
     ylab="Life Expectancy", main="Life expectancy as a function of population")
```
This plot shows a slight positive correlation overall for this data set. However, there are streaks of data which are likely all from the same country just with the growing population size overtime.  

## More data processing
Finding what years have NAs for infant mortality
```{r}
yrNA<-data.frame() #create an empty data frame that can be added on to later
#create a for loop to go through all years in the data set
#count the number of TRUE meaning there is an NA present in which year
#add year and number of NAs present in the empty data frame
for(i in 1960:2016){
  x<-sum(is.na(africadata$infant_mortality[which(africadata$year==i)]))
  yrNA <- rbind(yrNA,c(i,x))
}
```

Creating a new data frame for only the year 2000 in Africa
```{r}
africa2000 <- subset(africadata, year == "2000") #subsetting the data to only yr2000
summary(africa2000) #summary statistics of all variables
str(africa2000) #information on all the variables in the dataframe
```

## More plotting
Recreating previous plots with only data from the year 2000 in Africa
```{r}
#plot(x value, y value, x-axis label, y-axis label, main title)
plot(africa2000$infant_mortality, africa2000$life_expectancy, xlab="Infant Mortality",
     ylab="Life Expectancy", main="Life expectancy by infant mortality in 2000")
plot(log(africa2000$population), africa2000$life_expectancy, 
     xlab="Log of Population Size", ylab="Life Expectancy",
     main="Life expectancy by population in 2000")
```

## A simple fit
```{r}
#fitting a linear model to infant mortality and life expectancy
#lm(response~terms) - where terms means predictors
fit1 <- lm(africa2000$life_expectancy ~ africa2000$infant_mortality)
summary(fit1)

#fitting a linear model to population and life expectancy
fit2 <- lm(africa2000$life_expectancy ~ africa2000$population)
summary(fit2)
```
Predicting life expectancy based on infant mortality has a significant positive correlation. While using population size to predict life expectancy is not significantly correlated.
